D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-01-25 ootavacuous.tex
January 25 self-review #1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

@@@ This section provides a brief overview of the OOTA problem, followed
by an equally brief summary of prior work in this area, and ending
with a quick review of a few of the challenges for general definitions of
``OOTA cycle''.

@@

However, it is important to distinguish OOTA cycles from the simple
reordering exemplified by
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ implementation\footnote{
	This paper will use ``C++ implementation'' both excluding the
	CPU, as here, and including the CPU.
	Explicit disambiguation is added when necessary.}
and the CPU are within their rights to reorder lines~12 and~13, which
can result in all of \co{x}, \co{y}, \co{r1}, and \co{r2} having the
value 42 as follows:

@@

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been looked upon favorably by C++ implementers.

Goldblatt looked at interactions between OOTA cycles and
UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
Appendix~\ref{app:Aside on Undefined Behavior}
analyzes the examples and notes ways to separate UB and OOTA-cycle
concerns.
The remainder of this document will focus on examples lacking UB.

@@

Listing~\ref{lst:OOTA Cycle}
shows a code fragment that in strict C++ would have an OOTA cycle.
This section describes this fragment, thus giving an overview of
the \co{herd7} tool.

@@

Finally, line~23 specifies an \co{exists} clause, which gives a
condition to check for the final value of the specified local
variables.
The \co{0:} prefix specifies a variable local to \co{P0()} and
the \co{1:} prefix specifies a variable local to \co{P1()}.
The \co{/\\} specifies a boolean AND, and the \co{=} specifies
an equality comparison.
If a variable appears in an \co{exists} clause, then the final value of
that variable constitutes observable behavior.

@@

Listing~\ref{lst:OOTA Cycle, herd7 Output} shows the corresponding output
of the \co{herd7} tool.
Lines~3 and~4 show the possible states, with line~4 showing the
counterintuitive outcome where both processes load the value 42.
As we will see, this outcome is not possible in real-world implementations
due to physical constraints of which \co{herd7} is (by design) unaware.
Line~9 contains \co{Sometimes} (as opposed to \co{Never} or \co{Always}),
indicating that some executions satisfy the \co{exists} clause and
others do not.

@@

The standard's prohibition of OOTA is of course important, but those of us
writing code in the real world must rely on actual C++ implementations.
And in these implementations, this prohibition is in fact enforced by TSO
ordering in strongly ordered systems and by data dependency ordering in
weakly ordered systems.
Real-world implementations cannot store something until after they load
it, and this \emph{semantic dependency} results in compilers emitting
instructions such that the hardware enforcement applies.\footnote{
	The need to prohibit simple OOTA is one reason why compiler-based
	value speculation optimizations are frowned upon.}
Roughly speaking, there is semantic dependency from a given load to
a given store when a change in the value loaded that can change the
value stored.

@@

However, the ordering implied by the semantic dependency is only one
type of link in an OOTA cycle.
The other link extends from a given store to the load that returns the
value stored.
It is tempting to make the intuitive argument that this store must precede
this load in global time, and then to combine this with the notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that such implementations cannot exhibit
OOTA cycles.
This combination has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf, where sdep is the set of semantic dependencies within
each thread and rf is the set of store-to-load links, whether intra-thread
(rfi) or between threads (rfe).\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for definitions
	and properies of rf, rfe, and rfi.}

@@

This is a fine definition, and is consistent with the words in the C++
standard, except for code fragments exemplified by
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}.
This is an elaboration of
Listing~\ref{lst:Simple OOTA}
that adds \co{z} and adds lines~15 and~16 to \co{thread2()}.
The problem is that the C++ implementation is within its rights to
note that line~16 might well execute immediately after line~15, so that
no other thread had a chance to modify \co{z}.
Such an implementation might therefore behave as if the source code
had instead been as shown in
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized},
with line~16 optimized away in favor of line~17 storing the value
in \co{r2} that was loaded from \co{y}.
And given that the load from \co{z} no longer exists, it cannot possibly
act as a temporal constraint.

In order to avoid these backwards-in-time rfi links, we substitute rfe
for rf, thus defining an OOTA cycle as a cycle in sdep $\cup$ rfe (or
\co{acyclic(sdep ; rfe)} in \co{herd7} notation), where rfe is the set
of store-to-load links, but excluding intra-thread rfi links, which are
instead considered to be part of sdep.
Although this does shunt additional complexity onto the term
``semantic dependency'', it also enables us to cleanly separate
the inter-thread and intra-thread portions of any given OOTA cycle.
% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}
@@@ Please see
Appendix~\ref{app:Informal Definition of Semantic Dependency}
for more a more detailed (albeit quite informal) definition.

>>>>

This section provides a brief overview of the OOTA problem, followed
by an equally brief summary of prior work in this area, and ending
with a quick overview of the \co{herd7} tool that will be used to
evaluate litmus tests.

@@

However, it is important to distinguish OOTA cycles from the OOTA-like
behavior that can be caused by simple reordering.
An example of simple reordering is shown in
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ implementation\footnote{
	This paper will use ``C++ implementation'' both excluding the
	CPU, as here, and including the CPU.
	Explicit disambiguation is added when necessary.}
and the CPU are within their rights to reorder lines~12 and~13, which
can result in all of \co{x}, \co{y}, \co{r1}, and \co{r2} having the
value 42 as follows:

@@

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been looked upon favorably by C++ implementers.
Some of these workers appear to have abandoned this effort, but as of
early 2024, Mark Batty is persisting with modular relaxed dependencies.

Goldblatt looked at interactions between OOTA cycles and
UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
Appendix~\ref{app:Aside on Undefined Behavior}
analyzes his examples and notes ways to separate UB and OOTA-cycle
concerns.
The remainder of this document will focus on examples lacking UB.

@@

Listing~\ref{lst:OOTA Cycle}
shows a code fragment that in strict C++ would have an OOTA cycle (but
that is of course prohibited in full C++).
This section describes this fragment, thus giving an overview of
the \co{herd7} tool.

@@

Line~23 gives an example of a \co{locations} clause, which causes
\co{herd7} to dump out the final value of \co{x} and \co{y}.

Finally, line~24 specifies an \co{exists} clause, which gives a
condition to check for the final value of the specified local
variables.
The \co{0:} prefix specifies a variable local to \co{P0()} and
the \co{1:} prefix specifies a variable local to \co{P1()}.
The \co{/\\} specifies a boolean AND, and the \co{=} specifies
an equality comparison.
If a variable appears in an \co{exists} clause, then the final value of
that variable constitutes observable behavior.

@@

Listing~\ref{lst:OOTA Cycle, herd7 Output} shows the corresponding output
of the \co{herd7} tool.
Lines~3 and~4 show the possible states, with line~4 showing the
counterintuitive outcome where both processes load the value 42.
Normally these states would include only those variables mentioned
in the \co{exists} clause, but because of the \co{locations} clause
on line~23, the values of \co{x} and \co{y} is also included, which
can be helpful when debugging.
As we will see, this outcome is not possible in real-world implementations
due to physical constraints of which \co{herd7} is (by design) unaware.
Line~9 contains \co{Sometimes} (as opposed to \co{Never} or \co{Always}),
indicating that some executions satisfy the \co{exists} clause and
others do not.

@@

The standard's prohibition of OOTA is of course important, but those of us
writing code in the real world must rely on actual C++ implementations.
And in these implementations, this prohibition is in fact enforced by TSO
ordering in strongly ordered systems and by data dependency ordering in
weakly ordered systems.
Real-world implementations cannot store something until after they load
it, and this \emph{semantic dependency} results in compilers emitting
instructions such that the hardware enforcement applies.\footnote{
	The need to prohibit simple OOTA is one reason why compiler-based
	value speculation optimizations require checks on such
	speculation, and these checks must be based on actual values
	loaded.}
Roughly speaking, there is semantic dependency from a given load to
a given store when a change in the value loaded that can change the
value stored.

@@

However, semantic dependencies are only one type of link in an OOTA cycle.
The other link extends from a given store to the load that returns the
value stored.
It is tempting to make the intuitive argument that this store must precede
this load in global time, and then to combine this with the notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that such implementations cannot exhibit
OOTA cycles.
This combination has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf, where sdep is the set of semantic dependencies within
each thread and rf is the set of store-to-load links, whether intra-thread
(rfi) or between threads (rfe).\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for definitions
	and properies of rf, rfe, and rfi.}

@@

This is a fine definition, and is consistent with the words in the C++
standard, but has an intra-thread problem exemplified by
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}.
This is an elaboration of
Listing~\ref{lst:Simple OOTA}
that adds \co{z} and adds lines~15 and~16 to \co{thread2()}.
The problem is that the C++ implementation is within its rights to
note that line~16 might well execute immediately after line~15, giving
threads no chance to modify \co{z}.
Such an implementation might therefore behave as if the source code
had instead been as shown in
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized},
with line~16 optimized away in favor of line~17 storing the value
in \co{r2} that was loaded from \co{y}.
And given that the load from \co{z} no longer exists, it cannot possibly
act as a temporal constraint.

In order to avoid these rfi links, we substitute rfe
for rf, thus defining an OOTA cycle as a cycle in sdep $\cup$ rfe (or
\co{acyclic(sdep ; rfe)} in \co{herd7} notation), where rfe is the set
of store-to-load links, but excluding intra-thread rfi links, which are
instead considered to be part of sdep.
Although this does shunt additional complexity onto the term
``semantic dependency'', it also enables us to cleanly separate
the inter-thread and intra-thread portions of any given OOTA cycle.
% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}

@@@ Please see
Appendix~\ref{app:Informal Definition of Semantic Dependency}
for more a more detailed (albeit quite informal) definition.
Maybe eventually have an appendix that lists all the definitions
of both ``OOTA cycle'' and ``semantic dependency''.
